---
title: "Open data"
format: html
editor: visual
toc: true
toc-title: ""
toc-location: right
---

```{r, echo=F,eval=T,message=F,warning=F}
packages = c("ggplot2", "readxl", "dplyr", "tidyr", "xtable")
# Install packages not yet installed
installed_packages = packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}
# Packages loading
invisible(lapply(packages, library, character.only = TRUE))
```

## Roadmap

In this section, we are going to learn how to:

-   Open datasets
-   Perform simple data cleaning tasks
-   Merge datasets, and check the merge before believing it

## Open datasets

### Data format

As you may know, data come in various formats. Here are some common formats. The last column is about read speed and file size on large datasets, not about how good the format is in general.

| Extension  | Format  | Description            | Speed on large files |
|------------|---------|------------------------|----------------------|
| `.csv`     | CSV     | Comma-separated values | ⭐⭐                 |
| `.tsv`     | TSV     | Tab-separated values   | ⭐⭐                 |
| `.xlsx`    | XLSX    | Excel spreadsheet      | ⭐                   |
| `.RData`   | RData   | R's own format         | ⭐⭐⭐               |
| `.json`    | JSON    | Nested, web-oriented   | ⭐⭐                 |
| `.parquet` | PARQUET | Columnar storage       | ⭐⭐⭐⭐⭐           |

Many other formats exist, depending on the source or the type of data stored (eg, spatial data). Note that SQL is a database *system* rather than a file format, and you query it rather than open it, with the `DBI` package.

::: callout-note
Speed is not the only criterion, and usually not the main one. A `.csv` can be opened by anything, forever, including by a co-author who does not use R. A `.RData` file cannot. For anything you intend to share or archive, prefer `.csv`. For a 20 GB intermediate file that only your own code will ever read, use `.parquet`.
:::

### Open data in R

We are going to manipulate mostly `.csv` databases. They are non-proprietary and can be opened by virtually all software. This is not the most efficient way to store data. If you manipulate *very* large databases (tens of millions of observations with hundreds of columns), I advise you to use `.parquet`.

Base R includes functions to open `.csv` files: `read.csv()` and `read.csv2()`. Some packages might be required to open other formats. To open a large `.csv` dataset, I advise you to use the `fread()` function from the package `data.table`.

The syntax is:

```{r, eval = F}
data = read.csv2(file = "C:/your/path/data_to_open.csv")
```

::: callout-important
## The first silent failure of this course

`read.csv()` expects a **point** as the decimal separator and a **comma** as the field separator. `read.csv2()` expects a **comma** as the decimal separator and a **semicolon** as the field separator. French data, and INSEE data in particular, are usually the second kind.

Use the wrong one and R does not complain. It just reads your numeric column as text. Then `log(rent)` throws `non-numeric argument to binary operator` three scripts later, and you have no idea why.

**So: run `str(data)` immediately after every import.** Every column that should be a number must show `num` or `int`. If it shows `chr`, you used the wrong reader. This takes two seconds and it will save you an afternoon at least once this year.
:::

### Write data

Once you are done cleaning the data, you might want to save your dataset (this is actually good practice!). Most functions to read data have a writing counterpart. Hence:

```{r, eval = F}
write.csv2(data, file = "C:/your/path/name_of_the_dataset.csv", row.names = FALSE)
```

Note `row.names = FALSE`. Without it, R writes an extra unnamed column containing 1, 2, 3, ... and when you read the file back you get a mysterious column called `X`. Every R user has met it. Now you know where it comes from.

### Use a `paste()`

You might want to save a bit of time, reduce the probability of writing errors, and increase the reliability and reproducibility of your code by using a `path` object within a `paste()` function. This function, from base R, simply concatenates strings. For instance `paste("Hello", "world", sep = " ")` will produce `"Hello world"`. An alternative function is `paste0()`, which by default concatenates without a space. Hence `paste("Hello", "world", sep = "")` and `paste0("Hello", "world")` are strictly equivalent and print `Helloworld`.

It comes in handy with the `read.csv2()` and `write.csv2()` functions. Instead of rewriting the full path each time, you can call the `path` object you created earlier. If the dataset `idf_data.csv` is stored in the subfolder `/raw_data`, it is easy to open it:

```{r, eval = F}
path = "C:/Users/mateomoglia/Dropbox/courses/polytechnique/2027_eco51432ep"

data = read.csv2(file = paste0(path, "/raw_data/idf_data.csv"))
```

It is useful if you share code with someone. They need to change only the line with the path, instead of the whole code.

### Discover the data

I prepared a dataset, mainly drawn from INSEE data, and downloaded from the Institut Paris Region, a research organisation on urbanism in the Ile-de-France region. It contains various socio-economic variables at the municipality (and Paris arrondissement) level. It contains information on demographic characteristics, income, and housing market variables.

The codebook contains all variables and a quick description.

## Data cleaning

Data cleaning refers to going from a raw dataset to a dataset that can be used and manipulated effortlessly. Below are some examples of data cleaning procedures:

-   Format column names (eg, replace spaces with underscores with `rename()`, or clean the whole set at once with `janitor::clean_names()`)
-   Remove missing values if they are *real* missing values, with `mutate()` and `ifelse()`
-   Replace weird values by NAs or 0s (eg, most INSEE missing wage values are coded 99999 and should be treated as NAs, not as 99999) with `mutate()` and `ifelse()`
-   Clean or create ID variables (for instance, the municipal ID in France is made of the *département* code on 2 digits and the municipality on 3 digits, so to build a municipal ID you concatenate both) with `mutate()` and `paste0()`
-   Put variables in the right format (eg, numeric variables such as wages should be... numeric and not string, which is very frequent with `.csv` datasets that do not store the column format) with `mutate()` and `as.numeric()` or `as.character()`
-   Overall visual inspection, with `str()`, `summary()` and `head()`

Although time consuming, this step is particularly important to (i) familiarise yourself with the data and (ii) save precious time when it comes to the analysis.

::: callout-warning
## The 99999 problem, and why it matters more than it looks

If a wage variable codes missing values as 99999 and you do not catch it, nothing breaks. `mean()` returns a number. The number is wrong, the table is wrong, the regression is wrong, and every one of them looks completely normal.

This is the archetype of the silent failure. The defence is not cleverness, it is habit: after every import, run `summary()` on every numeric column and look at the min and the max. A maximum of 99999 on a wage variable, or a minimum of 0 on a variable that cannot be 0, is visible in one second if you look, and invisible forever if you do not.
:::

Usually, data come with a codebook, a dictionary, or some kind of documentation (another dataset or a `.pdf`). It is very useful to read it carefully to understand how data are reported (missing values, column names, variable formatting).

For the purpose of the class, the data are already cleaned, but never **take data at face value**!

## Data manipulation

### Merge datasets

In real world situations, data come from many sources and/or are delivered in many datasets (one per year, or per area, for instance). You then want to **merge** them to obtain a bigger and more insightful dataset.

::: callout-note
Because R is an object-based language, you can manipulate several datasets within the same function. However, this is perilous, as you have to make sure they have the same dimensions, are formatted the same way, and are sorted the same way. Merging on an explicit key removes all three risks at once.
:::

In `dplyr`, merging is done through the `join` functions. A join between `df1` and `df2` has to be done on at least one joining variable, called the key. There are several types of join:

| Join Type | Description |
|--------------------------|----------------------------------------------|
| `inner_join()` | Returns rows with matching keys in both tables |
| `left_join()` | Returns all rows from the left table, with matches from the right if present |
| `right_join()` | Returns all rows from the right table, with matches from the left if present |
| `full_join()` | Returns all rows from both tables, matching where possible |
| `semi_join()` | Returns rows from the left table that have a match in the right table |
| `anti_join()` | Returns rows from the left table that do **not** have a match in the right table |

The code looks like this. **Always write `by =` explicitly**, even when you do not have to:

```{r, echo = T, eval = F}
merged_df = left_join(df1, df2, by = "codgeo")
```

If the key does not have the same name in the two tables, name both sides:

```{r, echo = T, eval = F}
merged_df = left_join(df1, df2, by = c("codgeo" = "code_commune"))
```

Of course, this is compatible with the pipe syntax:

```{r, echo = T, eval = F}
merged_df = df1 %>%
  left_join(df2, by = "codgeo")
```

::: callout-important
## Never accept a join you have not checked

If you omit `by =`, `dplyr` guesses the key from every column name the two tables have in common. It prints a message saying so, which everyone ignores. If both tables happen to contain a column called `year`, or `libelle`, you have silently joined on two keys instead of one, and you have quietly lost most of your observations.

Three lines, every single time you merge:

```{r, echo = T, eval = F}
nrow(df1)                                            # rows you started with
nrow(merged_df)                                      # rows you ended with
df1 %>% anti_join(df2, by = "codgeo") %>% nrow()     # left rows with no match
```

For a `left_join`, `nrow(merged_df)` must equal `nrow(df1)`. If it is **larger**, your key is not unique in `df2` and rows have been duplicated. If the `anti_join` count is large, your keys do not have the same format, and 90% of the time the reason is a leading zero: `"01004"` read as the number `1004`. Check with `class(df1$codgeo)`. A municipal ID is text, never a number.

`anti_join()` is not an exotic verb you will never use. It is the diagnostic tool of this entire course.
:::

### Reshape datasets

Data manipulation also involves reshaping datasets. Let's say a dataset contains the population at the municipal level. The first column is the municipality ID and each other column is the population for a given year, such that: `codgeo`, `pop1999`, `pop2000`, up to `pop2023`.

To obtain a panel dataset (multiple IDs observed multiple times), we need to reshape it into **long format** (as opposed to wide). The function is `pivot_longer()`. It comes from the library `tidyr`. **You need to install it first, then load it**.

```{r, echo = T, eval = F}
data_long = data_wide %>%
  tidyr::pivot_longer(
    cols = starts_with("pop"), # Might have used -codgeo
    names_to = "year",         # New col name
    values_to = "population",  # New col name
    names_prefix = "pop"       # Remove the prefix 
  ) %>%
  mutate(year = as.integer(year)) # Not necessary, but usually year is numeric
```

The counterpart function is `tidyr::pivot_wider()`.

::: callout-tip
An easy way to remember which is which: **long**er makes the table longer (more rows, fewer columns), **wid**er makes it wider (more columns, fewer rows). After a pivot, check `dim()` before and after. The number of cells carrying information should not change.
:::

### Other manipulations

In `dplyr`, it is possible to manipulate data using the following functions:

| Function | Description |
|--------------------------|----------------------------------------------|
| `filter()` | Selects rows based on logical conditions (eg, keep Ile-de-France) |
| `select()` | Chooses columns by name |
| `arrange()` | Reorders rows based on column values |
| `distinct()` | Returns unique rows |
| `slice()` | Selects rows by position or helper functions (eg, `slice_head()`) |
| `bind_rows()` | Stack two dataframes by rows |
| `bind_cols()` | Combine two dataframes side by side |

::: callout-warning
`bind_cols()` glues two tables side by side **by position**, with no key and no check. If the two tables are not sorted identically you have just attached each municipality's rent to a different municipality's population, and nothing will tell you. Use a join. `bind_cols()` is for cases where you built both objects one line earlier and you are certain.
:::

::: callout-tip
## Exercise

1.  Open the population dataset `/raw_data/idf_pop.csv` and check whether it is clean. Run `str()` and `summary()` before anything else.
2.  Create a subset containing only `codgeo` and the columns that start with `population`
3.  Make it long, with `codgeo` as the ID and a `year` column
4.  Extract the most recent year (ideally with a function rather than by typing the year)
5.  From the same population file, create a second subset with `codgeo` and the seven CSP columns (`agri`, `cadre`, `profint`, `employ`, `ouvri`, `retraite`, `autsap`)
6.  Open the housing cost dataset `/raw_data/idf_rents.csv`
7.  Create a dataset with 3 columns: `codgeo`, `rent_appartement` and `rent_house`
8.  Merge the three datasets on the municipality ID. **Check the merge with the three lines above before you go further.**
9.  Save the merged dataset as `.csv` in `/temp`

This dataset will be our main dataset for the rest of the class, so it is worth getting right.
:::

::: {.callout-note collapse="true"}
## Solution

Column names may differ slightly in your file, check with `names()` and adapt.

```{r, echo = T, eval = F}
#-------------------------------------------------------------------------------
# 1_build_idf_data.R
# Builds the main dataset for the class: population, CSP, rents.
#
# Author: your name
# Date: today
#-------------------------------------------------------------------------------

library(dplyr)
library(tidyr)

path = "C:/Users/mateomoglia/Dropbox/courses/polytechnique/2027_eco51432ep"

# 1. Open and inspect ----------------------------------------------------------

pop = read.csv2(file = paste0(path, "/raw_data/idf_pop.csv"))

str(pop)      # every numeric column must be num or int, and codgeo must be chr
summary(pop)  # look at the min and the max of every column

# 2-4. Population in the most recent year --------------------------------------

pop_last = pop %>%
  select(codgeo, starts_with("population")) %>%
  pivot_longer(
    cols = starts_with("population"),
    names_to = "year",
    values_to = "population",
    names_prefix = "population"
  ) %>%
  mutate(year = as.integer(year)) %>%
  filter(year == max(year, na.rm = TRUE)) # a function, not a hardcoded 2021

# 5. Population by CSP ---------------------------------------------------------

pop_csp = pop %>%
  select(codgeo, agri, cadre, profint, employ, ouvri, retraite, autsap)

# 6-7. Rents -------------------------------------------------------------------

rents = read.csv2(file = paste0(path, "/raw_data/idf_rents.csv")) %>%
  select(codgeo, rent_appartement, rent_house)

# 8. Merge, and check ----------------------------------------------------------

idf_data = pop_last %>%
  left_join(pop_csp, by = "codgeo") %>%
  left_join(rents,   by = "codgeo")

nrow(pop_last)                                          # expected
nrow(idf_data)                                          # obtained: must be equal
pop_last %>% anti_join(rents, by = "codgeo") %>% nrow()  # municipalities with no rent

# 9. Save ----------------------------------------------------------------------

write.csv2(idf_data,
           file = paste0(path, "/temp/idf_data.csv"),
           row.names = FALSE)

sessionInfo()
```

If the `anti_join` returns a non-zero count, do not shrug. Look at the municipalities concerned. Some rents are genuinely missing for very small communes, which is fine and expected. But if the count is in the hundreds, your keys have a formatting problem, not a data problem.
:::

::: callout-note
## Before moving on, you should be able to

-   [ ] Say what `str()` tells you that `head()` does not
-   [ ] Explain in one sentence why `read.csv()` and `read.csv2()` are not interchangeable
-   [ ] Write the three lines that check a merge, from memory
-   [ ] Say how many rows a `left_join` should return, before running it
-   [ ] Find your saved `idf_data.csv` on disk without using the RStudio file pane
:::